---
output:
  html_document:
    toc: true
    theme: united
  knitrBootstrap::bootstrap_document:
    theme.chooser: TRUE
    highlight.chooser: TRUE
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Introduction to derfinder}
-->

Introduction to `derfinder`
===========================

```{r vignetteSetup, echo=FALSE, message=FALSE}
## Track time spent on making the vignette
startTime <- Sys.time()

## Bib setup
library('knitcitations')

## Load knitcitations with a clean bibliography
cleanbib()
cite_options(hyperlink = 'to.doc', citation_format = 'text', style = 'html')
# Note links won't show for now due to the following issue
# https://github.com/cboettig/knitcitations/issues/63

## Write bibliography information
write.bibtex(c(derfinder = citation('derfinder'), 
    knitrBootstrap = citation('knitrBootstrap'), 
    knitr = citation('knitr')[3],
    rmarkdown = citation('rmarkdown')),
    file = 'derfinderRef.bib')
bib <- read.bibtex('derfinderRef.bib')

## Fix some names to work with CRAN and GitHub versions
names(bib)[names(bib) == 'hester2013knitrbootstrap'] <- 'hester2014knitrbootstrap'
```


# Overview

This is a stub vignette.






# Reproducibility

Code for creating the vignette

```{r createVignette, eval=FALSE, bootstrap.show.code=FALSE}
## Create the vignette
library('knitrBootstrap') 

knitrBootstrapFlag <- packageVersion('knitrBootstrap') < '1.0.0'
if(knitrBootstrapFlag) {
    ## CRAN version
    library('knitrBootstrap')
    system.time(knit_bootstrap('derfinder.Rmd', chooser=c('boot', 'code'), show_code = TRUE))
    unlink('derfinder.md')
} else {
    ## GitHub version
    library('rmarkdown')
    system.time(render('derfinder.Rmd', 'knitrBootstrap::bootstrap_document'))
}
## Note: if you prefer the knitr version use:
# library('rmarkdown')
# system.time(render('derfinder.Rmd', 'html_document'))


## Extract the R code
library('knitr')
knit('derfinder.Rmd', tangle = TRUE)
```

Date the vignette was generated.

```{r reproducibility1, echo=FALSE, bootstrap.show.code=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproducibility2, echo=FALSE, bootstrap.show.code=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits=3)
```

`R` session information.

```{r reproducibility3, echo=FALSE, bootstrap.show.code=FALSE, bootstrap.show.message=FALSE}
## Session info
library('devtools')
session_info()
```

# Bibliography

This vignette was generated using `knitrBootstrap` `r citep(bib[['hester2014knitrbootstrap']])`
with `knitr` `r citep(bib[['xie2014knitr']])` and `rmarkdown` `r citep(bib[['allaire2014rmarkdown']])` running behind the scenes.

Citations made with `knitcitations` `r citep('https://github.com/cboettig/knitcitations')`.

```{r vignetteBiblio, results='asis', echo=FALSE}
## Print bibliography
bibliography()
```
